Safe Transferable Regions

### 2016-07-05 paper/implementation.tex
More comments from DV.
minor changes

We have implemented the prototype of \name compiler frontend,
including its region type system and type inference, in 3k+ lines of
OCaml. Our implementation is called \namec. The input to \namec is a
program in $\absof{\FB^+}$, an extended version of $\absof{\FB}$ that
includes assignments, conditionals, loops, more primitive datatypes
(\eg, integers), and a null value. 
% If needed, a specification file containing region type annotations
% for higher-order functions can also be provided. 
Our implementation of region type inference and constraint solving
closely follows the description given in
Sec.~\ref{sec:type-inference}. The one difference is that our
constraint solver uses an approximation algorithm to compute the
minimal graph augmentation (\S~\ref{sec:csolve}).
% To solve the
% constraints that arise during type inference, we built a solver called
% \csolve that implements constraint solving approach based on fixpoint
% computation and graph augmentation described in \S~\ref{sec:csolve}. 
% Since graph augmentation is NP-hard, \csolve uses an approximation
% algorithm that repeatedly adds an edge and recomputes paths in graph
% $G_1$ until it is as connected as graph $G_2$. Though this approach
% may not necessarily return the weakest solution to the abduction
% problem (\S~\ref{sec:csolve}), it seems to do so in practice, since
% constraints are not often complex. 
% If the input $\absof{\FB^+}$ program does not create any unsafe
% references, \namec annotates it with region types, which act as a
% witness to program's memory safety. If potential safety violations are
% encountered, then the type inference fails during the constraint
% solving phase.

@@

Our experience with Naiad benchmarks suggests that, although our
approach cannot statically enforce memory safety, it is nonetheless
useful, particularly because the static verification comes at no
additional cost to the developer. If the developer is satisfied with
the confidence gained by static verification, she may even choose to
turn the runtime safety checks off. Since we did not implement code
generation in \namec, we could not measure the runtime overhead of
checks needed to enforce safety. However, the number of LOC peforming
operations on \C{Region} objects relative to the total LOC never
exceeds 8\% in Naiad benchmarks. 
% \C{SelectVertex}, for example, contains one each of
% \C{open} and \C{transfer} operations on one region in its 45LOC,
% whereas, \C{JoinVertex} contains 5 \C{open}s, 2 \C{transfer}s, and 1
% \C{free} operation.

>>>>

We have implemented the prototype of \name compiler frontend,
including its region type system and type inference, in 3k+ lines of
OCaml. Our implementation is called \namec. The input to \namec is a
program in $\absof{\FB^+}$, an extended version of $\absof{\FB}$ that
includes assignments, conditionals, loops, more primitive datatypes
(\eg, integers), and a null value. 
% If needed, a specification file containing region type annotations
% for higher-order functions can also be provided. 
Our implementation of region type inference and constraint solving
closely follows the description given in
Sec.~\ref{sec:type-inference}.

% To solve the
% constraints that arise during type inference, we built a solver called
% \csolve that implements constraint solving approach based on fixpoint
% computation and graph augmentation described in \S~\ref{sec:csolve}. 
% Since graph augmentation is NP-hard, \csolve uses an approximation
% algorithm that repeatedly adds an edge and recomputes paths in graph
% $G_1$ until it is as connected as graph $G_2$. Though this approach
% may not necessarily return the weakest solution to the abduction
% problem (\S~\ref{sec:csolve}), it seems to do so in practice, since
% constraints are not often complex. 
% If the input $\absof{\FB^+}$ program does not create any unsafe
% references, \namec annotates it with region types, which act as a
% witness to program's memory safety. If potential safety violations are
% encountered, then the type inference fails during the constraint
% solving phase.

@@

Our experience with Naiad benchmarks shows the utility of our
type inference/checking tool,
%suggests that, although our
%approach cannot statically enforce memory safety, it is nonetheless
%useful, 
particularly because it comes at no additional cost to the developer.
If the developer is satisfied with
the confidence gained by static verification, she may even choose to
turn the runtime safety checks off. Since we did not implement code
generation in \namec, we could not measure the runtime overhead of
checks needed to enforce safety. However, the number of LOC peforming
operations on \C{Region} objects relative to the total LOC never
exceeds 8\% in Naiad benchmarks. 
% \C{SelectVertex}, for example, contains one each of
% \C{open} and \C{transfer} operations on one region in its 45LOC,
% whereas, \C{JoinVertex} contains 5 \C{open}s, 2 \C{transfer}s, and 1
% \C{free} operation.

### 2016-07-05 paper/related.tex
minor changes

Tofte and Talpin in~\cite{tofte93,tofte94,tofte97} introduce the
concept of a region type system to statically enforce the safety of
their region-based memory management scheme in ML. Following their
seminal work, static type systems for safe region-based memory
management have been extensively studied in the context of various
languages and problem settings~\cite{cyclone02, cyclone04, yates99,
MIT03, DPJ09, HMN01, WW01, rust, gpu14}. Our work differs from the
existing proposals in the following ways. Firstly, our problem setting
includes lexically scoped stack regions and dynamic transferable
regions (both programmer-managed) in context of an object-oriented
programming language equipped with higher-order functions. Second, we
adopt a two-pronged approach to memory safety that relies on a
combination of a simple static type discipline and lightweight runtime
checks. 
% The type system reduces the problem of ensuring safety of all memory
% accesses in a program into a problem of enforcing a state-transition
% discipline on memory accesses over a small subset of objects in the
% program. The later can be done efficiently at runtime. This lets us
% guarantee safety in presence of
In particular, our approach circumvents the need for restrictive
static mechanisms (e.g., linear types and unique pointers) and
expensive runtime mechanisms (e.g., garbage collection and reference
counting) in order to guarantee safety. Lastly, our region type system
comes equipped with full type inference that eliminates the need to
write region annotations on types to convince the type checker that
the program is safe.

@@

Our work differs from Cyclone also in terms of its technical
contributions. While Cyclone equips C with a range of region
constructs~\cite{cycloneSCP}, the semantics of (a significant subset
of) such constructs, and the safety guarantees of the language are not
formalized. In contrast, the (static and dynamic) semantics of Broom
has been rigorously defined with respect to a well-understood formal
system (FGJ). The safety guarantees have been formalized and proved.
The core of Broom is very simple; the rules that make up static and
dynamic semantics occupy less than a page each. We believe that the
rigor and simplicity of Broom makes it easy to understand the the
underlying ideas, and apply them in various problem settings. Similar
contrast can be made of region type inference in both the languages.
Cyclone’s type inference was only ever described as being similar to
Tofte and Talpin’s, and it’s effectiveness in presence of tracked
pointers is not clear. In contrast, the complete Ocaml (pseudo) code
of Broom's inference algorithm, was given in the supplement and the
ideas underlying type inference have been described elaborately in the
paper

>>>>

Tofte and Talpin in~\cite{tofte93,tofte94,tofte97} introduce the
concept of a region type system to statically enforce the safety of
their region-based memory management scheme in ML. Following their
seminal work, static type systems for safe region-based memory
management have been extensively studied in the context of various
languages and problem settings~\cite{cyclone02, cyclone04, yates99,
MIT03, DPJ09, HMN01, WW01, rust, gpu14}. Our work differs from the
existing proposals in the following ways. Firstly, our problem setting
includes lexically scoped stack regions and dynamic transferable
regions (both programmer-managed) in context of an object-oriented
programming language equipped with higher-order functions. Second, we
adopt a two-pronged approach to memory safety that relies on a
combination of a simple static type discipline and lightweight runtime
checks. 
% The type system reduces the problem of ensuring safety of all memory
% accesses in a program into a problem of enforcing a state-transition
% discipline on memory accesses over a small subset of objects in the
% program. The later can be done efficiently at runtime. This lets us
% guarantee safety in presence of
In particular, our approach circumvents the need for restrictive
static mechanisms (e.g., linear types and unique pointers) or
expensive runtime mechanisms (e.g., garbage collection and reference
counting) in order to guarantee safety. Lastly, our region type system
comes equipped with full type inference that eliminates the need to
write region annotations on types to convince the type checker that
the program is safe.

@@

Our work differs from Cyclone also in terms of its technical
contributions. While Cyclone equips C with a range of region
constructs~\cite{cycloneSCP}, the semantics of (a significant subset
of) such constructs, and the safety guarantees of the language are not
formalized. In contrast, the (static and dynamic) semantics of Broom
has been rigorously defined with respect to a well-understood formal
system (FGJ). The safety guarantees have been formalized and proved.
The core of Broom is very simple; the rules that make up static and
dynamic semantics occupy less than a page each. We believe that the
rigor and simplicity of Broom makes it easy to understand the the
underlying ideas, and apply them in various problem settings. Similar
contrast can be made of region type inference in both the languages.
Cyclone’s type inference was only ever described as being similar to
Tofte and Talpin’s, and it’s effectiveness in presence of tracked
pointers is not clear.
%In contrast, the complete Ocaml (pseudo) code
%of Broom's inference algorithm, was given in the supplement and the
%ideas underlying type inference have been described elaborately in the
%paper

### 2016-07-05 paper/type-inference.tex
minor changes

\paragraph{Constraint Generation.}
The constraint generation algorithm is a direct adaption of the type checker:
each type checking rule is modified to produce a set of constraints that must hold
for the type checker to succeed.

Fig.~\ref{fig:constraint-gen} illustrates this for selected language constructs.
These rules use the same context as the expression typing judgment in
Fig.~\ref{fig:fb-staticsem} with symbols $\A$,$\ralloc$, and $\env$
retaining their meaning.
The algorithm proceeds top-down, analyzes an expression $e$ in a context
$\exptycx{\ralloc}{\env}$, and returns a type $\tau$ and a set of constraints $C$
(expressed in the rules as $\exprok {\exptycx{\ralloc}{\env}} {e} {\tau} {C}$ ),
indicating that the expression will have a type $\tau$ provided the constraints $C$ hold.

@@

@@

\item Our technique for region parametrization also uses a heuristic in the case of
higher order programs. Consider a higher-order method that has a function $f$ as a
parameter. We have a choice in terms of where the region-parameters introduced for
$f$ are bound (i.e., quantified). We always bind these region-parameters within the type
of $f$ (essentially, requiring $f$ to be region-polymorphic). This may also be a potential
source of incompleteness in the type-inference.

The treatment of class fields with a function type also uses the same (incomplete)
heuristic.

>>>>

\paragraph{Constraint Generation.}
The constraint generation algorithm is a direct adaption of the type checker:
each type checking judgment is modified to produce a set of constraints that must hold
for the type checker to succeed.

Fig.~\ref{fig:constraint-gen} illustrates this for selected language constructs.
These rules use the same context as the corresponding typing judgment in
Fig.~\ref{fig:fb-staticsem} with symbols $\A$,$\ralloc$, and $\env$
retaining their meaning.
The algorithm proceeds top-down, analyzes an expression $e$ in a context
$\exptycx{\ralloc}{\env}$, and returns a type $\tau$ and a set of constraints $C$
(expressed in the rules as $\exprok {\exptycx{\ralloc}{\env}} {e} {\tau} {C}$ ),
indicating that the expression will have a type $\tau$ provided the constraints $C$ hold.

@@

One notable aspect here is the following. Consider a method with predicate
variable $\varphi_m$ in a class with predicate variable $\varphi_c$. 
We use $\varphi_m$ as the constraint formula in the initial context used for
the method (instead of $\varphi_m \conj \varphi_c$ as in the static semantics).
It is possible to show that there is no loss of generality in doing so.
Specifically, it can be shown for any well-typed program, replacing the constraint
$\varphi_m$ of every method by the constraint $\varphi_m \conj \varphi_c$,
where $\varphi_c$ is the constraint of the class containing the method,
produces a well-typed program.
%
This simplification produces a restricted form of constraints that is easier to solve.

@@

\item Our technique for region parametrization also uses a heuristic in the case of
higher order programs. The following examples illustrates that principal types may
not exist for higher order functions.
\begin{codejava}
unit apply ( T $\rightarrow$ unit f, T x, T y) {
   f(x); f(y);
}
\end{codejava}
This method may be typed assuming either that \C{f} is polymorphic over the region
that its parameter is allocated in (permitting \C{x} and \C{y} to be allocated in any
regions), or by assuming that \C{x} and \C{y} are allocated in the same region $\rho_1$
that \C{f} expects its parameters to be allocated in. Neither type subsumes the other.
Our algorithm heuristically chooses the first option, as it appears to be the more likely
and useful candidate.

%as either\\
%$
%\inang{ \rho_0 \rho_1 \rho_2}
%    (\inang{\rho_0 \rho_3} (T \inang{\rho_3} \xrightarrow{} unit, 
%      T \inang{\rho_1}, 
%      T \inang{\rho_2})
%   \xrightarrow{} 
%      unit
%$\\
%or as\\
%$
%\inang{ \rho_0 \rho_1}
%    (\inang{\rho_0} (T \inang{\rho_1} \xrightarrow{} unit, 
%      T \inang{\rho_1}, 
%      T \inang{\rho_1})
%   \xrightarrow{} 
%      unit
%$\\

%Consider a higher-order method that has a function $f$ as a
%parameter. We have a choice in terms of where the region-parameters introduced for
%$f$ are bound (i.e., quantified). We always bind these region-parameters within the type
%of $f$ (essentially, requiring $f$ to be region-polymorphic). This may also be a potential
%source of incompleteness in the type-inference.

% The treatment of class fields with a function type also uses the same (incomplete) heuristic.
